---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# facereaderconverter

<!-- badges: start -->
<!-- badges: end -->

The package is a series of functions for converting FaceReader output files into a more analysis-friendly format, and for detecting episodes of emotion from the time series data. It also includes some utilities for downstream analysis.

Tested to work with FaceReader 9.1.

## Installation

You can install the development version of facereaderconverter from [GitHub](https://github.com/) with:

```r
# install.packages("pak")
pak::pak("mmu-research-cloud-platform/facereaderconverter")
```
or

```r
# install.packages("devtools")
devtools::install_github("mmu-research-cloud-platform/facereaderconverter")
```

## File conversion

### `loadFRfile()`

`loadFRfile()` loads a FaceReader export into memory without writing an output file. It dispatches to the TXT, Excel, or CSV reader based on the file extension.

```{r load-fr-file-example, eval = FALSE}
library(facereaderconverter)

loadFRfile(
  inpath = "testdata/testdata_detailed.txt",
  values_as_numeric = TRUE,
  clean_names = TRUE
)
```

Use this when you want a parsed data frame directly from a FaceReader export.

### `convertFRFiles()`

`convertFRFiles()` reads a single FaceReader `.txt` file, skips the metadata header, guesses the delimiter, and writes a `.csv` with the same basename unless `return_data = TRUE`.

```{r convert-fr-files-example, eval = FALSE}
library(facereaderconverter)

convertFRFiles(
  inpath = "testdata/testdata_detailed.txt",
  values_as_numeric = TRUE,
  clean_names = TRUE
)
```

When `values_as_numeric = TRUE`, the `Video Time` column is converted to `hms` time and detailed-file values are converted to numeric where possible. When `clean_names = TRUE`, the output names are converted with `janitor::clean_names()`.

### `convertFRExcelFiles()`

`convertFRExcelFiles()` reads a single FaceReader `.xlsx` file, detects the header row automatically, and returns the parsed data unless `return_data = FALSE`.

```{r convert-fr-excel-example, eval = FALSE}
library(facereaderconverter)

convertFRExcelFiles(
  inpath = "FaceReaderOutput.xlsx",
  return_data = TRUE,
  values_as_numeric = TRUE,
  clean_names = TRUE
)
```

When `return_data = FALSE`, the function returns metadata about the workbook import.

### `convertFRDirectory()`

`convertFRDirectory()` processes all `.txt` files in a directory and returns metadata invisibly.

```{r convert-fr-directory-example, eval = FALSE}
library(facereaderconverter)

convertFRDirectory(
  inpath = "testdata",
  outpath = "junk",
  values_as_numeric = TRUE,
  cores = 2L
)
```

Useful arguments include `recursive` for searching subdirectories, `pattern` for filtering file names, `metadata_filename` for the output metadata file, and `duplicate_timecodes_as_error` for handling repeated time codes. The metadata output includes success and failure status columns.

## Episode coding

### `convert_to_episodes()`

`convert_to_episodes()` converts FaceReader coding data into detected episodes using hysteresis thresholds, a delta-based reaction signal, and a minimum duration filter. It accepts long data with `id`, `subject`, and either `video_time` or `frame`, plus `emotion` and `value`; if `emotion` and `value` are missing, wide data are reshaped to long format internally.

The function requires `id` and `subject`. If `frame` is not supplied, it is derived from `video_time` using `fps`. `T_up` and `T_down` control episode entry and exit, while `delta` and `delta_window` define the returned delta-up signal rather than episode boundaries.

Important arguments:

- `fps`: sampling rate in frames per second
- `T_up`: threshold for entering an episode
- `T_down`: threshold for leaving an episode
- `delta`: minimum change required by the delta rule
- `delta_window`: window size in seconds for the delta rule
- `min_dur_sec`: minimum episode duration in seconds
- `consecutive_missing`: maximum allowed run of missing values while an episode is active
- `cores`: number of threads used by `data.table`

The function returns an `fr_coding` object with four components: `episodes`, `deltas`, `coding`, and `metadata`. `episodes` contains the detected episode spans, `deltas` contains delta-up reaction events, and `coding` contains the annotated frame-level data with episode and delta markers.

```{r convert-to-episodes-example, eval = FALSE}
library(facereaderconverter)

coding_df <- read.csv("testdata/testdata_detailed.csv") |>
  dplyr::mutate(id = 1, subject = "parent")

coding_df2 <- coding_df |>
  tidyr::pivot_longer(
    cols = c(neutral, happy, sad, angry, surprised, scared, disgusted),
    names_to = "emotion",
    values_to = "value"
  )

res <- convert_to_episodes(
  coding_df2,
  fps = 30L,
  T_up = 0.20,
  T_down = 0.18,
  delta = 0.10,
  delta_window = 0.1,
  min_dur_sec = 0.1,
  consecutive_missing = 150L
)

res$episodes
res$deltas
res$coding
```

Episodes are grouped within each `id`, `subject`, and `emotion` combination. Episode end frames are adjusted to the last in-state frame with a non-missing `value`, and episodes shorter than the minimum duration are removed. Delta rows are tracked separately for downstream functions such as `reaction_rate()`.


## Downstream analysis



### `synchrony()`

`synchrony()` compares the episode coding for one subject against another subject within each `id` and emotion. The `missing_threshold` argument controls how much missing comparison data is tolerated within a denominator episode before that episode is dropped.

```{r synchrony-example, eval = FALSE}
library(facereaderconverter)

coding <- data.table::data.table(
  id = rep(1L, 6),
  subject = rep(c("teen", "parent"), each = 3),
  emotion = "happy",
  video_time = rep(1:3, 2),
  value = c(0.1, 0.2, 0.3, 0.1, 0.2, 0.3),
  in_state = c(FALSE, TRUE, FALSE, FALSE, FALSE, TRUE),
  run_id = c(1L, 1L, 1L, 2L, 2L, 2L)
)

episodes <- data.table::data.table(
  id = 1L,
  subject = c("teen", "parent"),
  emotion = "happy",
  run_id = c(1L, 2L),
  start_frame = c(2L, 3L),
  end_frame = c(2L, 3L)
)

coded_data <- structure(
  list(coding = coding, episodes = episodes),
  class = c("fr_coding", "list")
)

synchrony(coded_data, missing_threshold = 0.5)
```

The result reports denominator and numerator subjects, the emotion, the number of episodes, and synchrony. The `id` column keeps the same type as the input data.

### `synchrony_by_episode()`

`synchrony_by_episode()` returns the episode-level comparison table used by `synchrony()`. It includes one row per denominator episode, the numerator subject matched to that episode, and a logical `synchrony` flag indicating whether the comparison subject was present during the denominator episode.

```{r synchrony-by-episode-example, eval = FALSE}
library(facereaderconverter)

synchrony_by_episode(coded_data, missing_threshold = 0.5)
```

Use this when you need the underlying episode-level matches rather than the aggregated synchrony summary.

### `reaction_rate()`

`reaction_rate()` estimates the proportion of eligible episodes that contain at least one delta reaction after an initial exclusion window.

```{r reaction-rate-example, eval = FALSE}
library(facereaderconverter)

coded_data <- data.table::data.table(
  id = rep(1L, 8),
  subject = rep(c("teen", "parent"), each = 4),
  emotion = "happy",
  frame = rep(1:4, 2),
  delta = c(1L, 1L, 0L, 1L, 0L, 1L, 1L, 0L),
  in_state = rep(c(TRUE, TRUE, FALSE, FALSE), 2),
  run_id = c(1L, 1L, NA, NA, 2L, 2L, NA, NA)
)

reaction_rate(coded_data, fps = 30)
```

This is useful when you want a simple summary of reaction frequency by subject and emotion.

### `reaction_rate_by_episode()`

`reaction_rate_by_episode()` returns the episode-level table used by `reaction_rate()`. It includes one row per denominator episode with the matched numerator subject, the episode window, the number of valid frames, and a logical `reaction` flag.

```{r reaction-rate-by-episode-example, eval = FALSE}
library(facereaderconverter)

reaction_rate_by_episode(coded_data, fps = 30)
```

Use this when you need to inspect which specific episodes contributed to the aggregated reaction-rate summary.


### `locf()`

`locf()` applies last-observation-carried-forward logic to an `fr_coding` object and returns the imputed coding table plus episode summaries.

```{r locf-example, eval = FALSE}
library(facereaderconverter)

coding_df <- read.csv("testdata/testdata_detailed.csv") |>
  dplyr::mutate(id = 1, subject = "parent")

coding_df2 <- coding_df |>
  tidyr::pivot_longer(
    cols = c(neutral, happy, sad, angry, surprised, scared, disgusted),
    names_to = "emotion",
    values_to = "value"
  )

fr <- convert_to_episodes(coding_df2, fps = 30L)
locf(fr)
```

Use this when short missing stretches should inherit the most recent non-missing run assignment.

## Utilities

### `to_seconds()`

`to_seconds()` converts FaceReader-style timestamps such as `hh:mm:ss` or `hh:mm:ss.mmm` into numeric seconds.

```{r to-seconds-example}
library(facereaderconverter)

c(
  to_seconds("00:00:10", digits = 0L),
  to_seconds("00:00:10.500"),
  to_seconds("00:01:00", digits = 0L)
)
```

Use this when you need a numeric time variable for plotting or joining.

### `parse_time_to_frame()`

`parse_time_to_frame()` converts timestamps to frame indices at a given sampling rate. It accepts `HH:MM:SS`, `MM:SS`, or plain seconds.

```{r parse-time-to-frame-example}
library(facereaderconverter)

parse_time_to_frame(c("00:00:10.5", "1:23.5", "83.5"), fps = 30)
```

This is the helper used internally when `video_time` is present but `frame` is not.

### `map_paths()`

`map_paths()` remaps files from one root directory to another while preserving the relative folder structure.

```{r map-paths-example, eval = FALSE}
library(facereaderconverter)

map_paths(
  input_dir = "data/raw",
  output_dir = "data/converted",
  files = c(
    "data/raw/session1/a.txt",
    "data/raw/session2/b.txt"
  )
)
```

This is useful when converting a directory of files into a parallel output tree.

## Exported function dependencies

Most exported functions are independent helpers, but several depend on other exported functions or on custom structures produced by other functions. A direct call is not always the same as a workflow dependency, so both are listed separately.

| Exported function | Direct exported-function dependencies | Required structure/class or input shape | Common upstream producer(s) |
|---|---|---|---|
| `loadFRfile()` | `convertFRFiles()`, `convertFRExcelFiles()` | none | none |
| `convertFRDirectory()` | `convertFRFiles()`, `map_paths()` | none | none |
| `convert_to_episodes()` | `parse_time_to_frame()` | creates `fr_coding` | none |
| `add_delta_column()` | `parse_time_to_frame()` | accepts `fr_coding`; otherwise needs a coding table with `id`, `subject`, `emotion`, `frame` or `video_time`, and `value` | `convert_to_episodes()` |
| `delta_episodes()` | none | needs `delta` plus `id`, `subject`, `emotion`, `frame` (and optionally `video_time`, `value`) | `add_delta_column()` |
| `locf()` | none | accepts `fr_coding`; otherwise needs a coding table with `id`, `subject`, `emotion`, `video_time`, `value`, and `run_id` | `convert_to_episodes()` |
| `reaction_rate()` | none | accepts `fr_coding`; otherwise needs a delta-coded table with `id`, `subject`, `emotion`, `frame`, and `delta` | `add_delta_column()` |
| `reaction_rate_by_episode()` | none | accepts `fr_coding`; otherwise needs a delta-coded table with `id`, `subject`, `emotion`, `frame`, and `delta` | `add_delta_column()` |
| `synchrony()` | none | requires `fr_coding` returned by `convert_to_episodes()` | `convert_to_episodes()` |
| `synchrony_by_episode()` | none | requires `fr_coding` returned by `convert_to_episodes()` | `convert_to_episodes()` |
| `convertFRFiles()` | none | none | none |
| `convertFRExcelFiles()` | none | none | none |
| `to_seconds()` | none | none | none |
| `parse_time_to_frame()` | none | none | none |
| `map_paths()` | none | none | none |
